Phylogram instead of chronogram when assessing the neutral evolution of a trait

### 2025-08-14 manuscript/main-first-submission.tex
Added manuscript tex file and prepare for revision

% \title{Phylograms disentangle drift and selection acting on a continuous trait.}
% \title{Drift and selection acting on a continuous trait can be disentangled by a simultaneous use of a chronogram and a phylogram}
% \title{Phylogram instead of chronogram when claiming that a trait is evolving under drift.}
\title{Phylogram for testing neutral evolution of a trait, chronogram to model selection.}
\rhead{\scshape Chronogram or phylogram for trait evolution.}

\author{
\large
\textbf{T. {Latrille}$^{1}$\orcidlink{0000-0002-9643-4668}, T. {Gaboriau}$^{1}$\orcidlink{0000-0001-7530-2204}, N. {Salamin}$^{1}$\orcidlink{0000-0002-3963-4954}}\\
\normalsize
$^{1}$Department of Computational Biology, Université de Lausanne, Lausanne, Switzerland\\
\texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
}
\begin{document}
%TC:ignore

\begin{center}
    \Huge Cover letter
\end{center}
{\Large
Dear Editor,\\

We are pleased to submit our manuscript entitled ”Phylogram for testing neutral evolution of a trait, chronogram to model selection” for consideration for publication in Evolution Letters.\\

In this study, we revisit a puzzling question of phylogenetic comparative methods: does the time that have passed since the species diverged explain differences in traits? Or is it the genetic differences between species that explains best the differences? While the de facto standard in PCM is to use the time to explain the differences in traits, this reasoning is not always justified. Indeed, if the trait is evolving neutrally, we argue that the genetic differences between species instead of time should explain best the differences in traits. As a result, in this study, we show and argue that using simultaneously both the genetic differences and the time since species diverged provides a better understanding of how traits evolve across species.\\

We believe that our manuscript will be of great interest to the readers of Evolution Letters as it offers a novel perspective on trait evolution and provides practical guidelines for researchers in the field. We confirm that this manuscript has not been published elsewhere and is not under consideration by another journal.\\

With best regards,\\

Thibault Latrille, Thibault Gaboriau \& Nicolas Salamin
}
\newpage
\maketitle

% Abstract (≤ 300 words)
\begin{abstract}
At the species level, the evolution of traits is driven by a combination of selective and neutral forces. To disentangle these processes, different scenarios of evolution are modelled and compared. For traits under selection, the species is seen as tracking a trait optimum, and such optimum can change along the species tree. On the other hand, neutral evolution is modelled as the trait changing randomly along the different branches of the species tree. Regardless of the intricacy of modelling trait changes, the species tree is assumed to be known and obtained independently, and the standard is to use a chronogram, where the branch lengths are proportional to time. Rationally, the time that has passed correlates with trait changes, even under a neutral regime, through the direct proxy of the number of generations that occurred. However, we argue that since species generation time can also vary along the phylogenetic tree, chronograms introduces biases. Instead, for a phylogram, the branch lengths represent sequence divergence (i.e. the number of substitutions) which absorb the effect of changing generation time. Based on this observation, in this study, we first show using simulations that for a trait evolving neutrally the fit of a random evolution has more support on a phylogram than on a chronogram. However, while doing model comparison and testing different scenarios of selection, using a phylogram also leads to incorrect predictions. Given these results, we argue that we should use phylograms instead of chronograms when claiming that a trait is evolving under drift. Nevertheless, we support the fact that we should generally continue to use chronograms to model selection acting on a trait.
\end{abstract}

@@

\newpage
%TC:endignore
\section*{Lay summary}\label{sec:summary}
When we look at different species, we can see that some traits are different between them.
For example, the peak size of Darwin's finches varies between species.
The question we ask is: does the species divergence time explain the differences in traits?
Or is it the genetic differences between the species that best explains the differences?
The standard in evolutionary biology is to use time to explain the differences in traits.
However, if the trait is not under natural selection and it is instead changing due to accumulating mutations, the genetic differences between species instead of time should best explain the differences in traits.
As a result, we argue that using simultaneously both the genetic differences and the species divergence time provides a better understanding of how traits evolve across species.

\section*{Introduction}\label{sec:introduction}
% The model of neutral trait evolution, explained in the context of phylogenetic comparative methods.
By observing the changes in a trait along a lineage, it is possible to speculate on the regime of selection acting on it.
But even without selection, a trait might still change due to random genetic drift.
As such, distinguishing selection acting on a trait from neutral evolution (i.e. drift) requires determining which regime of evolution is supported by the observed changes.
At a larger scale, from the observed variations of traits across species, regimes of evolution are typically assessed using phylogenetic comparative methods, where traits are modelled as evolving along the branches of the species tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a, harmon_phylogenetic_2018}.
For example, to model neutral evolution, the mean trait value is said to follow a Brownian motion (BM), branching and evolving independently after each speciation event~\citep{felsenstein_phylogenies_1985, lynch_phenotypic_1986, hansen_translating_1996}.
In other words, for each branch of the tree, the value at the descendant node is normally distributed around the ancestral value, with a variance proportional to the branch length.
In this framework, reconstructing trait variation along the whole phylogeny as a BM can thus constitute a null model of neutral trait evolution.

% Alternative models of trait evolution and their pitfalls.
Alternatively to a simple BM, a trend in the BM is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
More complex models to detect selection have been proposed, notably the Ornstein-Uhlenbeck (OU) processes, where trait variation is constrained around an optimum value, which is often interpreted as a signature of stabilizing selection~\citep{hansen_stabilizing_1997, butler_phylogenetic_2004, beaulieu_modeling_2012}.
Methodologically, this alternative model of evolution raises issues since an OU process might be statistically preferred over a BM due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Alleviating statistical artifact and adding biological complexity and realism, the OU process can be relaxed to allow for multiple optima along the phylogenetic tree, which is interpreted as a few abrupt changes in the environment along the phylogeny~\citep{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020, grabowski_cautionary_2023}.
Finally, the optimum can also be allowed to change continuously along the tree, which results in the trait itself reflecting the movement of the optimum along the lineages due to constant adaptive evolution of the trait toward the optimum~\citep{hansen_translating_1996, hansen_three_2024}.
One special case of a continuously changing optimum is again the BM, where changes in optimum are the consequences of a randomly changing environment~\citep{hansen_comparative_2008}, alternatively named fluctuating selection~\citep{holstad_evolvability_2024} or also diversifying selection due to the diversity of generated phenotypes at the clade level~\citep{latrille_detecting_2024}.
Regardless of this discrepancy in terminology, this modelling raises an issue: the BM can not be interpreted as the null model of neutral trait evolution since it also models the alternative scenario of a continuously moving optimum.

% Deriving the neutral expectation from standing variation
In order to disentangle neutral evolution from selection, another approach is to contrast the observed rate of evolution to the neutral expectation~\citep{lande_genetic_1980}.
The neutral expectation can be obtained if the underlying genetic architecture of the trait is known and the trait encoded by many loci of additive effects~\citep{barton_infinitesimal_2017, sella_thinking_2019}.
Theoretically, can also relate changes in mean trait value along a lineage to the trait variance at the population scale~\citep{turelli_heritable_1984, felsenstein_phylogenies_1988a}.
Generalized to many species at the phylogenetic scale, contrasting between and within-species variations allow to disentangle neutral evolution from selection~\citep{latrille_detecting_2024}.
And more generally, comparing the rate of evolution at different timescales allow to infer the regime of evolution~\citep{hansen_three_2024, holstad_evolvability_2024}.
However, trait variation at different timescales is not always available, and the genetic architecture of the trait is often unknown.
Instead, to disentangle neutral evolution from selection in the context of phylogenetic comparative methods, we hereby focus on a blind spot of the framework: the underlying tree and the unit of its branch lengths (Fig.~\ref{fig:methods}A).

%TC:ignore
\begin{figure*}[!htb]
    \centering
    \includegraphics[width=\textwidth, page=1] {figures/artwork_rational}
    \caption{
        Panel~A: Across many species, the evolution of continuous traits can be modelled as a stochastic process evolving along the branches of a phylogenetic tree.
        The branches of such phylogenetic tree can be measured in either time (chronogram) or number of substitutions (phylogram).
        Panel~B: Theoretically, changes of a trait under a moving optimum should be better predicted by a chronogram.
        Instead, changes of a neutral trait should be better predicted a phylogram.
    }
    \label{fig:methods}
\end{figure*}
%TC:endignore

% The use of chronogram is widely accepted and de facto standard in phylogenetic comparative methods.
Typically, the tree is assumed to be known and obtained independently, and the de facto standard in phylogenetic comparative methods is to use a chronogram, where the branch lengths are proportional to time~\citep{felsenstein_phylogenies_1985, harmon_phylogenetic_2018}.
Theoretically, for a neutrally evolving trait, trait changes depend directly on the number of generations~\citep{hansen_translating_1996}, which is proportional to time only if the average time between two consecutive generations, called generation time, is constant and the same for each species.
Since generation time depends on the time for an individual to reach sexual maturity, it vary considerably between species (e.g. from a few months for coral reef pygmy gobies to 150 years for Greenland shark).
As a result, variations of generation time between species means that modelling neutral evolution as a BM on a chronogram might induce biases~\citep{litsios_effects_2012}.

% Phylogram to model neutral trait evolution
Alternatively, on a phylogram, the branch lengths represent another quantity than time, which can be used as the backbone to model trait evolution and can potentially absorb changes in generation time.
For example, it is known that using phylograms in units of morphological distance can be used to improve ancestral trait reconstruction for a discrete character~\citep{wilson_chronogram_2022}, but in this case the phylogram is not expected to absorb changes in generation time.
From a genomic perspective, phylograms can also be in units of nucleotide divergence, that is, depicting the number of nucleotide substitutions occurring along a branch.
Because the number of neutral nucleotide substitutions accumulating along a branch is proportional to the number of generations~\citep{kimura_evolutionary_1968, kimura_neutral_1983}, nucleotide divergence would in this case absorb the effect of changing generation time (i.e. longer branches for lineages with shorter generation time).
Such a phylogram, if obtained from neutrally evolving genomic loci would be more appropriate to model a trait evolving neutrally~\citep{latrille_detecting_2024}.
Additionally, the rational to use nucleotide divergence to absorb changes in generation time is also valid for changes in mutation rate (per loci and per generation) under the assumption of a constant genetic architecture of the trait and that changes in mutation rate impact the whole genome~\citep{latrille_detecting_2024}.
Empirically, such changes in mutation rate along a phylogenetic tree is also observed, although to a lesser extent than changes in generation time~\citep{bergeron_evolution_2023}.

% Phylogram in the case of selection
Under the alternative scenario of selection, shorter generation time and higher mutation rate can also allow for the species to track the changing optimum faster and not lag behind~\citep{lande_natural_1976, hansen_comparative_2008}.
However, the timescale for this lag (hundred or thousands of years) is negligible compared to timescale on which the optimum changes at the level of clades (millions of years) in first approximation~\citep{hansen_three_2024}.
Thus, at the phylogenetic scale, since shifts in the optimum fitness peak is extrinsic to the species and is dependent on the environment which varies with time, using a chronogram would be more accurate to model mean trait changes.
As a result, for a trait under selection, mean trait changes should be better explained by the time rather than the number of substitutions that occurred along a branch.

% Here, we propose a method to evaluate the soundness of studying trait evolution on a phylogram or a chronogram.
In this study, we test if using a phylogram (in units of nucleotide divergence for neutral loci) instead of a chronogram can allow to assess the regime of evolution of a trait (Fig.~\ref{fig:methods}B), by evaluating by the fit of a BM on both trees.
More precisely, we seek to test if the changes of a neutral trait are better predicted by a phylogram.
Conversely, we also seek to test if changes of a trait under a moving optimum are better predicted by a chronogram.
Using genomic information we seek to provide additional approaches to model trait evolution, particularly to disentangle selection and neutral evolution.

\section*{Methods}
In order to assess whether a phylogram is better suited to model neutral evolution we first performed simulations of a trait evolving along a phylogenetic tree under different regimes of selection (neutral, moving optimum, multiple optima).
Second, we gathered an empirical dataset of body and brain masses from mammals, including both a chronogram and a phylogram (in unit of nucleotide divergence from neutral loci).
Third, we fitted a single rate Brownian motion (BM) to the simulated data, and we compared the fit of the BM on a phylogram versus a chronogram.
Fourth, based in the single rate BM, we derived a model to estimate the support of a phylogram over a chronogram that we apply to the simulated and empirical datasets.
Finally, we fitted several models of trait evolution to the simulated and empirical datasets: a multi-rate BM, an Ornstein-Uhlenbeck (OU) process, and a relaxed OU with multiple optima.

\subsection*{Simulations along a phylogenetic tree}
We performed simulations under different selective regimes (neutral, moving optimum, multiple optima).
Simulations were individual-based and followed a Wright-Fisher model with mutation, selection and drift for a diploid population including speciation along a predefined ultrametric phylogenetic tree.
We used the same simulation framework as in \citet{latrille_detecting_2024}, with parameters detailed in the supplementary material (section~\ref{sec:simulator}).
The parameters of simulations were chosen to mimic an empirical dataset of mammals.
In summary, the trait was encoded by $\NbrLoci$ independent loci, with each locus contributing additively, and mutations were drawn from a Poisson distribution at each generation (Fig.~\ref{fig:results-brownian}A).
Parents were selected for reproduction according to their phenotypic value, with a probability proportional to their fitness.
Flattening the fitness landscape resulted in neutral evolution (Fig.~\ref{fig:results-brownian}B), which meant that each individual had the same probability of being sampled at each generation regardless of its trait value.
Alternatively, for a trait under selection around a moving optimum (Fig.~\ref{fig:results-brownian}B), we modelled stabilizing selection acting on the trait, with the optimum value changing as a geometric Brownian motion (BM) along the phylogenetic tree~\citep{hansen_stabilizing_1997, hansen_translating_1996}.
Finally, for a trait under selection around multiple optima~\citep{uyeda_novel_2014}, we draw a Bernoulli variable to test whether there is a switch of optimum along a branch; if a switch occurs, the sliding of the optimum is drawn from a reflected exponential distribution (symmetric positive and negative values).

At each node of the tree, the population is split into two daughter populations running independently on each of the two branches, and the process is repeated until the tips of the tree are reached (Fig.~\ref{fig:results-brownian}C).
As a control, we performed simulations with constant mutation rates, generation times and effective population sizes ($\Ne$).
Alternatively, we performed simulations with fluctuating mutation rates, generation times and $\Ne$, where we used a BM to model the long-term changes along the phylogenetic tree, and we overlaid short-term changes in $\Ne$ (see supplementary material section~\ref{sec:simulator}).

The phylogram is obtained from the same simulation setting on a set of independent 30000 neutral loci, and the nucleotide divergence is computed as the number of substitutions along each branch.
The chronogram is derived from the phylogram by fitting a relaxed molecular clock model (correlated rate model with penalized likelihood, default value) to the phylogram with the \textit{R} package \textit{ape}~\citep{paradis_ape_2004}.

@@

\section*{Funding}
Université de Lausanne

>>>>

\title{Phylogram instead of chronogram when assessing the neutral evolution of a trait}
\rhead{\scshape Phylogram over chronogram for neutral evolution.}

\author{
\large
\textbf{T. {Latrille}$^{1}$\orcidlink{0000-0002-9643-4668}, T. {Gaboriau}$^{1}$\orcidlink{0000-0001-7530-2204}, N. {Salamin}$^{1}$\orcidlink{0000-0002-3963-4954}}\\
\normalsize
$^{1}$Department of Computational Biology, University of Lausanne, Quartier Sorge, 1015 Lausanne, Switzerland\\
\texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
}
\begin{document}
%TC:ignore
\maketitle

% Abstract (≤ 300 words)
\begin{abstract}
At the species level, the evolution of traits is driven by a combination of selective and neutral forces. To disentangle these processes, different scenarios of evolution are modelled and compared. For quantitative traits under selection, the species is usually considered to track a trait optimum, and such an optimum can change along the branches of the species tree. On the other hand, neutral evolution is modelled with a trait changing randomly around the ancestral trait value along the different branches of the species tree. Regardless of the intricacy of modelling trait changes, the species tree is assumed to be known and obtained independently of the trait data analysed. Branch lengths are usually assumed to be in units proportional to time, and the tree is represented by a chronogram. The rationale is that time correlates with trait changes because of its direct connection with the number of generations that occurred. However, since the generation time of species can also vary along the phylogenetic tree, we argue that their use introduces biases. In contrast, if the phylogenetic tree is represented by a phylogram with branch lengths in units of sequence divergence, this will account for the effect of changing generation time. In this study, we show using simulations that, for a trait evolving neutrally, the fit of a random evolution model has more support on a phylogram than on a chronogram. However, comparing models and testing different scenarios of selection using a phylogram leads to incorrect predictions. Given these results, we argue that we should use phylograms instead of chronograms when claiming that a trait is evolving under drift. Nevertheless, we support the fact that we should generally continue to use chronograms to model selection acting on a quantitative trait.
\end{abstract}

@@

\newpage
%TC:endignore
\section*{Lay summary}\label{sec:summary}
When we look at different species, we can see that some traits are different between them.
For example, the brain size of different primate and human lineages varies.
The question we ask is: does the time that passed explain the differences in their traits?
Or are genetic differences between species better at explaining the differences?
The standard in evolutionary biology is to use time to explain differences in traits.
However, if the trait is not under natural selection and instead is changing due to accumulating mutations, the genetic differences between species instead of time should best explain the differences in traits.
As a result, we argue that using both genetic differences and species divergence time simultaneously provides a better understanding of how traits evolve across species.

\section*{Introduction}\label{sec:introduction}
% The model of neutral trait evolution, explained in the context of phylogenetic comparative methods.
By measuring changes in a phenotypic trait along a lineage, it is possible to infer the type of selection acting on it or whether this change is only due to random genetic drift.
At a larger scale, from the observed variations of traits across species, regimes of evolution are typically assessed using phylogenetic comparative methods, where traits are modelled as evolving along the branches of the species tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a, harmon_phylogenetic_2018}.
For example, to model neutral evolution, the mean trait value is said to follow a Brownian motion (BM), branching and evolving independently after each speciation event~\citep{felsenstein_phylogenies_1985, lynch_phenotypic_1986, hansen_translating_1996}.
In other words, for each branch of the tree, the value at the descendant node is normally distributed around the ancestral value, with a variance proportional to the branch length.
In this framework, reconstructing trait variation along the whole phylogeny as a BM can thus constitute a null model of neutral trait evolution.

% Alternative models of trait evolution and their pitfalls.
Alternatively to a simple BM, adding a trend parameter in the BM is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
More complex models to detect selection have been proposed, notably the Ornstein-Uhlenbeck (OU) processes, where trait variation is constrained around an optimum value, which is often interpreted as a signature of stabilizing selection~\citep{hansen_stabilizing_1997, butler_phylogenetic_2004, beaulieu_modeling_2012}.
Methodologically, this alternative model of evolution raises issues since an OU process might be statistically preferred over a BM due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Adding biological realism, the OU process can be relaxed to allow for multiple optima along the phylogenetic tree, which is interpreted as a few abrupt changes in the environment along the phylogeny~\citep{ingram_surface_2013, uyeda_novel_2014, khabbazian_fast_2016, mitov_fast_2020, grabowski_cautionary_2023}.
Finally, the optimum can also be allowed to change continuously along the tree, which results in the trait itself reflecting the movement of the optimum along the lineages due to the constant adaptive evolution of the trait toward the optimum~\citep{hansen_translating_1996, hansen_three_2024}.
One special case of a continuously changing optimum is again the BM, where changes in the optimum are a consequence of the randomly changing environment~\citep{hansen_comparative_2008}, named fluctuating selection~\citep{holstad_evolvability_2024} or also diversifying selection due to the diversity of generated phenotypes at the clade level~\citep{latrille_detecting_2024}.
This modelling raises an issue in disentangling selection from neutral evolution: the best fit of BM can not be interpreted as the trait evolving neutrally, it can also be interpreted as a continuously moving optimum.

% Deriving the neutral expectation from standing variation
In order to disentangle neutral evolution from selection, another approach is to contrast the observed rate of evolution to the neutral expectation~\citep{lande_genetic_1980}.
The neutral expectation can be obtained if the underlying genetic architecture of the trait is known and the trait encoded by many loci of additive effects~\citep{barton_infinitesimal_2017, sella_thinking_2019}.
Additionally, we can also derive the expected changes in mean trait value along a lineage given the trait variance at the population scale~\citep{turelli_heritable_1984, felsenstein_phylogenies_1988a}.
When generalized to many species at the phylogenetic scale, contrasting between and within-species variations allow us to disentangle neutral evolution from selection~\citep{latrille_detecting_2024}.
More generally, comparing the rate of evolution at different timescales allows inferring the regime of evolution~\citep{hansen_three_2024, holstad_evolvability_2024}.
However, trait variation at different timescales is not always available, and the genetic architecture of the trait is often unknown.
Instead, to disentangle neutral evolution from selection in the context of phylogenetic comparative methods, we hereby focus on the underlying tree and the unit of its branch lengths (Fig.~\ref{fig:methods}A).

%TC:ignore
\begin{figure*}[!htb]
    \centering
    \includegraphics[width=\textwidth, page=1] {figures/artwork_rational}
    \caption{
        Panel~A: Across many species, the evolution of continuous traits can be modelled as a stochastic process evolving along the branches of a phylogenetic tree.
        The branches of such phylogenetic tree can be measured in either time (chronogram) or number of substitutions (phylogram).
        Panel~B: Theoretically, changes of a trait under a moving optimum should be better predicted by a chronogram.
        Instead, changes of a neutral trait should be better predicted by a phylogram.
    }
    \label{fig:methods}
\end{figure*}
%TC:endignore

% The use of chronogram is widely accepted and de facto standard in phylogenetic comparative methods.
Typically, the tree is assumed to be known and obtained independently, and the \textit{de facto} standard in phylogenetic comparative methods is to use a chronogram, where the branch lengths are proportional to time~\citep{felsenstein_phylogenies_1985, harmon_phylogenetic_2018}.
Theoretically, for a neutrally evolving trait, trait changes depend directly on the number of generations~\citep{hansen_translating_1996}, which is proportional to time only if the average time between two consecutive generations, called generation time, is constant and equal for all species.
Since generation time depends on the time for an individual to reach sexual maturity, it can vary considerably between species, from a few months for coral reef pygmy gobies to 150 years for the Greenland shark~\citep{demagalhaes_database_2009, nielsen_eye_2016}.
As a result, variations of generation time between species means that modelling neutral evolution as a BM on a chronogram might induce biases~\citep{litsios_effects_2012}.

% Phylogram to model neutral trait evolution
Alternatively, on a phylogram, the branch lengths represent another quantity than time, which can be used as the backbone to model trait evolution and can potentially absorb changes in generation time.
For example, it is known that using phylograms in units of morphological distance can be used to improve ancestral trait reconstruction for a discrete character~\citep{wilson_chronogram_2022}.
From a genomic perspective, phylograms can also be in units of nucleotide divergence, that is, depicting the number of nucleotide substitutions occurring along a branch.
Because the number of neutral nucleotide substitutions accumulating along a branch is proportional to the number of generations~\citep{kimura_evolutionary_1968, kimura_neutral_1983}, nucleotide divergence would in this case absorb the effect of changing generation time (i.e. longer branches for lineages with shorter generation time).
Such a phylogram, if obtained from neutrally evolving genomic loci would be more appropriate to model a trait evolving neutrally~\citep{latrille_detecting_2024}.
Additionally, the rational to use nucleotide divergence to absorb changes in generation time is also valid for changes in mutation rate (per loci and per generation) under the assumption of a constant genetic architecture of the trait and that changes in mutation rate impact the whole genome~\citep{latrille_detecting_2024}.
Empirically, such changes in mutation rate along a phylogenetic tree are also observed, although to a lesser extent than changes in generation time~\citep{bergeron_evolution_2023}.

% Phylogram in the case of selection
Under the alternative scenario involving selection, shorter generation time and higher mutation rate can also allow for the species to track the changing optimum faster and not lag behind~\citep{lande_natural_1976, hansen_comparative_2008}.
However, the timescale for this lag (hundreds or thousands of years) is negligible compared to the timescale on which the optimum changes at the level of clades (millions of years) in a first approximation~\citep{hansen_three_2024}.
Thus, at the phylogenetic scale, since shifts in the optimum fitness peak is extrinsic to the species and is dependent on the environment which varies with time, using a chronogram would be more accurate to model mean trait changes.
As a result, for a trait under selection, mean trait changes should be better explained by the time rather than the number of substitutions that occurred along a branch (Fig.~\ref{fig:methods}B).

% Here, we propose a method to evaluate the soundness of studying trait evolution on a phylogram or a chronogram.
In this study, we test this hypothesis, namely if using a phylogram (in units of nucleotide divergence for neutral loci) instead of a chronogram (in units proportional to time) can allow to assess the regime of evolution of a trait, by evaluating the fit of a BM on both trees.
More precisely, we seek to test if the changes of a neutral trait are better predicted by a phylogram.
Conversely, we also seek to test if changes of a trait under a moving optimum are better predicted by a chronogram.
Using genomic information we seek to provide additional approaches to model trait evolution, particularly to disentangle selection and neutral evolution.

\section*{Methods}
To assess whether a phylogram is better suited to model neutral evolution we first performed simulations of a trait evolving along a phylogenetic tree under different regimes of selection (neutral, moving optimum, multiple optima).
Second, we fitted a single rate Brownian motion (BM) to the simulated data, and we compared the fit of the BM on a phylogram versus a chronogram.
Third, based on the single rate BM, we derived a model to estimate the support of a phylogram over a chronogram that we apply to the simulated dataset.
Fourth, we fitted several models of trait evolution to the simulated datasets: a multi-rate BM, an Ornstein-Uhlenbeck (OU) process, and a relaxed OU with multiple optima, tested on a phylogram versus a chronogram.
Finally, we gathered an empirical dataset of body and brain masses from mammals, including both a chronogram and a phylogram (in units of nucleotide divergence from neutral loci), on which we tested the support of a phylogram for the single rate BM.

\subsection*{Simulations along a phylogenetic tree}
We performed simulations under different selective regimes (neutral, moving optimum, multiple optima).
Simulations were individual-based and followed a Wright-Fisher model with mutation, selection and drift for a diploid population including speciation along a predefined ultrametric phylogenetic tree.
We used the same simulation framework as in \citet{latrille_detecting_2024}, with parameters detailed in the supplementary material (section~\ref{sec:simulator}).
The parameters of simulations were chosen to mimic an empirical dataset of mammals.
In summary, the trait was encoded by $\NbrLoci$ independent loci, with each locus contributing additively, and mutations were drawn from a Poisson distribution at each generation.
Parents were selected for reproduction according to their phenotypic value, with a probability proportional to their fitness.
Flattening the fitness landscape resulted in neutral evolution, which meant that each individual had the same probability of being sampled at each generation regardless of its trait value.
Alternatively, for a trait under selection around a moving optimum, we modelled stabilizing selection acting on the trait, with the optimum value changing as a geometric Brownian motion (BM) along the phylogenetic tree~\citep{hansen_stabilizing_1997, hansen_translating_1996}.
Finally, for a trait under selection around multiple optima~\citep{uyeda_novel_2014}, we draw a Bernoulli variable to test whether there is a switch of optimum along a branch; if a switch occurs, the sliding of the optimum is drawn from a reflected exponential distribution (symmetric positive and negative values).

At each internal node of the tree (i.e. speciation event), the population is split into two daughter populations running independently on each of the two branches, and the process is repeated until the tips of the tree are reached.
As a control, we performed simulations with constant mutation rates, generation times and effective population sizes ($\Ne$).
Alternatively, we performed simulations with fluctuating mutation rates, generation times and $\Ne$, where we used a BM to model the long-term changes along the phylogenetic tree, and we overlaid short-term changes in $\Ne$ (see supplementary material section~\ref{sec:simulator}).

The phylogram is obtained from the same simulation setting on a set of 30,000 independent neutral loci, and the nucleotide divergence is computed as the number of substitutions along each branch.
The chronogram is derived from the phylogram by fitting a relaxed molecular clock model (correlated rate model with penalized likelihood, default value) to the phylogram with the \textit{R} package \textit{ape}~\citep{paradis_ape_2004}.

\subsection*{Brownian motion (BM)}
On the simulated dataset, we fitted a single rate BM using either a phylogram or a chronogram as the underlying tree.
We used \textit{RevBayes}~\citep{hohna_revbayes_2016} to fit a Brownian motion (BM) on continuous traits evolving along the branches of a phylogenetic tree~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a}.
The data consist of the mean trait value for each extant species, and the tree topology is fixed.
Along each branch, the value of the trait is drawn from a normal distribution with mean equal to the parent node value and variance equal to the branch length~\citep{felsenstein_phylogenies_1985}.
Formally, the BM process is described by the equation:
% Ornstein-Uhlenbeck model equations
\begin{align}
    \der \Trait_t & = \sigma \Multiply \der W_t,  \label{eq:bm-process}
\end{align}
where $\Trait_t$ is the trait value at time $t$, $\sigma$ is the rate of the BM, and $\der W_t$ is a Wiener process (i.e. a standard Brownian motion).
We used a log-uniform prior for $\sigma$.
Using Markov chain Monte Carlo (MCMC), we reconstructed the ancestral trait value at each node of the tree as the posterior mean estimate (burn-in of 1000 gen., running of 10,000 gen., 2 chains), and we compared the accuracy of the reconstruction using either a phylogram or a chronogram as the underlying tree.
Both trees are scaled such that the sum of all the branch lengths is equal to one in each case.
Importantly, the data and the tree topology are the same in both analyses; only the branch lengths are different.
In this simulation setting, for traits simulated under neutral evolution, we expect more accurate trait reconstruction on a phylogram, and conversely, for traits under selection (moving optimum), more accurate trait reconstruction on a chronogram.

\subsection*{BM with a switch}
To test the support of a phylogram over a chronogram, we implemented a model based on a single rate BM.
The model was implemented in \textit{RevBayes} and contains a switch variable, denoted as $\PhyloSupport$ that allows the model to switch the branch lengths in units of time to units of substitutions (Fig.~\ref{fig:results-brownian}A).
Formally, $\PhyloSupport \in {0, 1}$ is a Bernoulli random variable ($\PhyloSupport \in {0, 1}$) with a prior probability of $0.5$.
If $\PhyloSupport$ is $0$, the branch lengths are those of the chronogram, and if $\PhyloSupport$ is $1$, the branch lengths are those of the phylogram.
The tree topology is fixed, meaning both trees have the same branching structure, but because branch lengths are different and not necessarily on the same scale, both trees are re-scaled by dividing branch lengths by the total tree length.
As in the previous section, a BM is fitted to the data by modelling trait changes as normal distributions (\textit{dnNormal} in \textit{RevBayes}) running along the branches of the tree, with variance proportional to the branch length and the squared value of the Brownian rate parameter ($\sigma$, log-uniform prior).
Alternatively for large trees or for faster computation, the likelihood can also be estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}).
Altogether, the input data is the mean trait value for extant species and both the chronogram and phylogram with the same tree topology.
The posterior mean of $\PhyloSupport$ (burn-in of 1,000 gen., running of 10,000 gen., 2 chains) is the probability that the phylogram is favoured over the chronogram.
We expect $\PhyloSupport = 1$ for a trait under neutral evolution, and $\PhyloSupport = 0$ for a trait under selection (moving optimum).

\subsection*{Alternative models of evolution}
In addition to a single rate BM, we also fitted a BM with multiple rate parameters~\citep{eastman_novel_2011}, using either a phylogram or a chronogram as the underlying tree to the simulated dataset.
The likelihood of the data is estimated by the REML method (\textit{dnPhyloBrownianREML} in \textit{RevBayes}), and the number of rate shifts is estimated by reversible-jump MCMC (\textit{dnReversibleJumpMixture} in \textit{RevBayes}).
For each branch, we draw a rate-multiplier either equal to $1$ (no rate shift), or drawn from a log-normal distribution with a median of 1, and a standard deviation such that rate shifts range over about one order of magnitude.
We set our prior to the expected number of rate shifts of $1$ across the whole tree.
The number of rate shifts as well as the variance of rate parameters are estimated as the posterior mean (burn-in of 1,000 gen., running of 50,000 gen., 2 chains).

% Simple OU model versus a brownian process.
Also, we compared the fit of an Ornstein-Uhlenbeck (OU) process (eq.~\ref{eq:ou-process}) to a BM~\citep{hansen_stabilizing_1997,butler_phylogenetic_2004}, using either a phylogram or a chronogram as the underlying tree.
% Ornstein-Uhlenbeck model equations
The OU process is described by the equation:
\begin{align}
    \der \Trait_t & = -\alpha \left( \Trait_t - \theta \right) \der t + \sigma \Multiply \der W_t, \label{eq:ou-process}
\end{align}
where the parameter $\alpha$ in eq.~\ref{eq:ou-process} is the strength of the pull towards the optimum, $\theta$ is the optimum value and the other parameters are defined as in eq~\ref{eq:bm-process}.
The likelihood of the data is estimated by the REML method (\textit{dnPhyloOrnsteinUhlenbeckREML} in \textit{RevBayes}), and because the OU process has more parameters than the BM, we used a reversible-jump MCMC switch between the two models ($0$ for BM and $1$ for OU, \textit{dnReversibleJumpMixture} in \textit{RevBayes}).
% alpha ~ dnReversibleJumpMixture(0.0, dnExponential( abs(root_age / 2.0 / ln(2.0)) ), 0.5)
The prior for $\theta$ is uniform between the minimum and maximum trait value for extant species.
The prior for $\alpha$ is a reversible-jump mixture distribution: a value of $0$ or drawn from an exponential distribution with mean equal to half the root age divided by $\ln (2)$, meaning that we expect a phylogenetic half life of half the tree age.
Thus, when $\alpha = 0$ the OU process (eq.~\ref{eq:ou-process}) is equivalent to a BM (eq.~\ref{eq:bm-process}).
The support for the OU model is estimated as the posterior mean of the switch variable ($p_{\text{OU}}$), or equivalently that of $\alpha \neq 0$ (burn-in of 1,000 gen., running of 10,000 gen., 2 chains).

% Relaxed OU model, how many optimums?
Finally, we fitted a relaxed OU with multiple optima~\citep{uyeda_novel_2014}, using either a phylogram or a chronogram as the underlying tree.
The likelihood of the data given the tree is estimated by the REML method (\textit{dnPhyloOrnsteinUhlenbeckREML} in \textit{RevBayes}), and the number of optimums is also estimated by the reversible-jump MCMC method (\textit{dnReversibleJumpMixture} in \textit{RevBayes}) as the posterior mean estimate (burn-in of 1,000 gen., running of 5,000 gen., 2 chains).
For each branch, we draw a shift in optimum either equal to $0$ (no shift), or drawn from a uniform distribution centred on $0$ with breadth spanning the whole range of observed trait value.
We set our prior to the expected number of optimum shifts of $1$ across the whole tree.

@@

\section*{Funding}
This work was funded by Faculté de Biologie et de Médecine, Université de Lausanne (https://www.unil.ch; to TL, TG and NS) and Swiss National Science Fund (https://www.snf.ch; grant 315230-219757 to TL and NS). The funders did not play any role in the study design, data collection and analysis, decision to publish, or preparation of the manuscript.
